### 2010-03-18 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: Some minor fixes and notes

In dependently-typed language, $\Sigma$-types can be interpreted in
two different ways. This duality is actually reflected in the notation
we can find in the literature, depending on the sensibility of the
author. Hence, the type $\SigmaTy{A}{B}$ is sometimes written
$\Sigma_{x : A} (B\: x)$. This notation stresses the fact that
$\Sigma$-types are a generalisation of sums over arbitrary
arities. When simply-typed languages have finite sums,
dependently-typed languages have sums indexed by any set.

On the other hand, $\SigmaTy{A}{B}$ is sometimes written
$\SIGMA{\V{x}}{A} (B\: x)$. Under this perspective, $\Sigma$-types can
be read as a generalisation of products, where the second component
can depend on the first one. When simply-typed languages pack data
into non-dependent tuples, dependently-types languages tolerate this
non-dependent usage. However, they also give the rather novel ability
for data to influence further data.

@@

To model these data-types, we simply need to capture this grammar in
our dependently-typed setting. Here, the notion of sum-of-product
naturally translates into \emph{sigmas-of-sigmas}. However, beyond
this intuition, it should be clear that a direct encoding of
data-types through raw $\Sigma$-types is not a viable option. Indeed,
just as for finite sets, encoding throw away information when we
crucially need it. In the realm of data-types, things become even more
tougher, as one could legitimately expect an induction principle, for
instance. Anonymous in a see of $\Sigma$s, our data-types have little
chance of survival. \note{Ridiculous tentative of poetry. I was
  really tired when I wrote that.}

@@

For our sigmas-of-sigmas to be known, we crucially need the ability to
name them. \note{Pursuing this crappy analogy. Solving the type
  presentation/representation debate would help solving that point.}
To this end, we will use a standard technique in dependently-typed
programming: we construct an ad-hoc universe. This technique dates
back to Martin-L\"of definition of type
theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
programming technique~\cite{benke:universe-generic-prog, whoelse?}
\note{I would need some other example of universe construction for
  practical purposes}. We refer the reader to the Agda
tutorial~\cite{norell:agda-tutorial} for a pedagogical presentation of
universe construction. \note{Is that really a motivating motivation
  for the usage of universe construction? }

@@

\[
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DUnit          & \Desc \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DIndx          & \Desc \To \Desc
\end{array}
}
&
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}ll@{\:=\:\:}ll}
\descop{\DUnit}{& X}        &  \Unit                                       \\
\descop{\DSigma{S}{D}}{& X} &  \SIGMAS{\V{s} : S}{\descop{D~s}{X}}         \\
\descop{\DIndx{D}}{& X}     &  \TIMES{X}{\descop{D}{X}}
\end{array}
}
\end{array}
\]

@@

@@

At this stage, we have recovered the data-types we are usually able to
define in simply-typed languages.

@@

Whereas this definition is amply sufficient in the simply-typed world,
it comes short of its promises in the dependently-typed realm. Indeed,
this type is not dependent. As this operator will be the corner stone
of any inductive definition, we are better off sharping it before use.

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\All{}{}{}{} : & \PITEL{\V{D}}{\Desc}
                 \PITEL{\V{X}}{\Set}
                 \PITEL{\V{P}}{\V{X} \To \Set} \\
               & \PI{\V{xs}}{\descop{D}{\Mu{X}}} 
                 \Set 
\end{array} \\
\All{\DId}{X}{P}{x} \mapsto P\: x \\
\All{(\DConst{Z})}{X}{P}{x} \mapsto Z \\
\All{(\DProd{D}{D'})}{X}{P}{\pair{d}{d'}{}} \mapsto \TIMES{\All{D}{X}{P}{d}}{\All{D'}{X}{P}{d'}} \\
\All{(\DSigma{S}{T})}{X}{P}{\pair{a}{b}{}} \mapsto \All{(T\: a)}{X}{P}{b} \\
\All{(\DPi{S}{T})}{X}{P}{f} \mapsto \PI{\V{s}}{S} \All{(T\: a)}{X}{P}{(f\: s)}
}
&
%%
\stk{
\begin{array}{@{}ll}
\all{}{}{}{}{} : & \PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X}}{\Set}
                   \PITEL{\V{P}}{\V{X} \To \Set} \\
                 & \PITEL{\V{R}}{\PI{\V{x}}{X} P\: x}
                   \PI{\V{xs}}{\descop{D}{\Mu{X}}} 
                   \All{D}{X}{P}{xs} 
\end{array} \\
\all{\DId}{X}{P}{R}{x} = R x \\
\all{(\DConst{Z})}{X}{P}{R}{x} = z \\
\all{(\DProd{D}{D'})}{X}{P}{R}{\pair{d}{d'}{}} = \pair{\all{D}{X}{P}{R}{d}}{\all{D'}{X}{P}{R}{d'}}{} \\
\all{(\DSigma{S}{T})}{X}{P}{R}{\pair{a}{b}{}} = \all{(T\: a)}{X}{P}{R}{b} \\
\all{(\DPi{S}{T})}{X}{P}{R}{f} = \LAM{s} \all{(T\: s)}{X}{P}{R}{(f\: s)}
\end{array}
}
\]

>>>>

In dependently-typed language, $\Sigma$-types can be interpreted in
two different ways. This duality is actually reflected in the notation
we can find in the literature, depending on the sensibility of the author.
The notation $\Sigma_{x : A} (B\: x)$ stresses the interpretation that
$\Sigma$-types are a generalisation of sums over arbitrary arities.  Where
simply-typed languages have finite sums, dependently-typed languages have
sums indexed by any set.

On the other hand, our choice of notation ($\SIGMA{\V{x}}{A} (B\: x)$)
emphasises that $\Sigma$-types can be read as a generalisation of products,
where the second component can depend on the first one. When simply-typed
languages pack data into non-dependent tuples, dependently-types languages
tolerate this non-dependent usage. However, they also give the rather novel
ability for data to influence types of further data.

@@

To model these data-types, we simply need to capture this grammar in
our dependently-typed setting. Here, the notion of sum-of-product
naturally translates into \emph{sigmas-of-sigmas}. However, beyond
this intuition, it should be clear that a direct encoding of
data-types through raw $\Sigma$-types is not a viable option. Indeed,
just as for finite sets, encoding throw away information when we
crucially need it. In the realm of data-types, things become even more
tougher, as one could legitimately expect an induction principle, for
instance. Anonymous in a sea of $\Sigma$s, our data-types have little
chance of survival. 
\note{Ridiculous tentative of poetry. I was really tired when I wrote that.}

@@

For the structure of our data-types to be known, we crucially need the
ability to name them. \note{Pursuing this crappy analogy. Solving the type
presentation/representation debate would help solving that point.} To this
end, we will use a standard technique in dependently-typed programming: we
construct an ad-hoc universe. This technique dates back to Martin-L\"of
definition of type theory~\cite{martin-lof:itt}. Since then, it has been
fruitful as a programming 
technique~\cite{benke:universe-generic-prog,whoelse?}
\note{I would need some other example of universe construction for
  practical purposes}. 
We refer the reader to the Agda tutorial~\cite{norell:agda-tutorial} for a
pedagogical presentation of universe construction. 
\note{Is that really a motivating motivation for the usage of universe 
construction?}
\note{pwm: Surely ($\EnumU$, $\EnumT$) is a useful example of universes too.
Even if it doesn't show off their full power?}

@@

\begin{figure}

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DUnit          & \Desc \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DIndx          & \Desc \To \Desc
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}ll@{\:=\:\:}ll}
\descop{\DUnit}{& X}        &  \Unit                                       \\
\descop{\DSigma{S}{D}}{& X} &  \SIGMAS{\V{s} : S}{\descop{D~s}{X}}         \\
\descop{\DIndx{D}}{& X}     &  \TIMES{X}{\descop{D}{X}}
\end{array}
}
\end{array}
}\]

@@

\end{figure}

@@

At this stage, we have recovered the data-types we are usually able to
define in simply-typed languages.
\note{pwm: Err, not quite --- not til we introduce $\DHindx{\!}{\!}$}

@@

Whereas this definition is amply sufficient in the simply-typed world,
it comes short of its promises in the dependently-typed realm. Indeed,
this type is not dependent. As this operator will be the corner stone
of any inductive definition, we are better off sharpening it before use.

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\All{}{}{}{} : & \PITEL{\V{D}}{\Desc}
                 \PITEL{\V{X}}{\Set}
                 \PITEL{\V{P}}{\V{X} \To \Set} \\
               & \PI{\V{xs}}{\descop{D}{\V{X}}} 
                 \Set 
\end{array} \\
\All{\DId}{X}{P}{x} \mapsto P\: x \\
\All{(\DConst{Z})}{X}{P}{x} \mapsto Z \\
\All{(\DProd{D}{D'})}{X}{P}{\pair{d}{d'}{}} \mapsto \TIMES{\All{D}{X}{P}{d}}{\All{D'}{X}{P}{d'}} \\
\All{(\DSigma{S}{T})}{X}{P}{\pair{a}{b}{}} \mapsto \All{(T\: a)}{X}{P}{b} \\
\All{(\DPi{S}{T})}{X}{P}{f} \mapsto \PI{\V{s}}{S} \All{(T\: a)}{X}{P}{(f\: s)}
}
&
%%
\stk{
\begin{array}{@{}ll}
\all{}{}{}{}{} : & \PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X}}{\Set}
                   \PITEL{\V{P}}{\V{X} \To \Set} \\
                 & \PITEL{\V{R}}{\PI{\V{x}}{X} P\: x}
                   \PI{\V{xs}}{\descop{D}{\V{X}}} 
                   \All{D}{X}{P}{xs} 
\end{array} \\
\all{\DId}{X}{P}{R}{x} = R x \\
\all{(\DConst{Z})}{X}{P}{R}{x} = z \\
\all{(\DProd{D}{D'})}{X}{P}{R}{\pair{d}{d'}{}} = \pair{\all{D}{X}{P}{R}{d}}{\all{D'}{X}{P}{R}{d'}}{} \\
\all{(\DSigma{S}{T})}{X}{P}{R}{\pair{a}{b}{}} = \all{(T\: a)}{X}{P}{R}{b} \\
\all{(\DPi{S}{T})}{X}{P}{R}{f} = \LAM{s} \all{(T\: s)}{X}{P}{R}{(f\: s)}
\end{array}
}
\]

### 2010-03-18 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: uniformize spacing in arrays
Desc@ICFP: uniformize EnumT definitions
Desc@ICFP: Some minor fixes and notes

Let us examine the consequences of our act. First of all, we discover
that the type theory does not actually need to be extended with the
type former $\EnumU$, the constructors $\NilE$ and $\ConsE$. We have
just defined $\EnumU$ above, by fix-point over the signature functor
$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
$\ListCons$ constructors of lists.

@@

We shall now fulfil our promises for $\Desc$. We are going to
implement the codes of the universe of descriptions. Interestingly, in
and by itself, a code is nothing but a data-type. We are in the same
situation than with $\EnumU$: we ought to be able to describe the
codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
first-class citizen, born with the standard, generic equipment of
data-types.

@@

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT [ \DUnit, \DSigma{}{}, \DIndx{} ])  \\
                           & \left[\begin{array}{l}
                                   \DUnit                                \\
                                   \DSigma{\Set}{(\LAM{\V{S}} ???)}      \\
                                   \DIndx{\DUnit}                        \\
                                   \end{array}
                             \right]
\end{array}
}\]

@@

\begin{figure*}

\[
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \ldots          & \:\:\ldots \\
    \DHindx         & \PI{H}{Set} \Desc \To \Desc
\end{array}
}
&
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\ldots                        &  \ldots \\
\descop{\DHindx{H}{D}}{X}     &  \TIMES{(H \To X)}{\descop{D}{X}}
\end{array}
}
\end{array}
\]

@@

\end{figure*}

@@

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT [ \DUnit, \DSigma{}{}, \DIndx{}, \DHindx{}{} ]) \\
                           & \left[\begin{array}{l}
                                   \DUnit                                            \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{S}{\DUnit})}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set}{(\LAM{\V{H}} \DIndx{\DUnit})}
                                   \end{array}
                             \right]
\end{array}
}\]

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT [ \ldots ]$
construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $f$ is pushed into the type
$\EnumT{e} \To \Desc$. This corresponds to a finite function
definition, as formalised in Figure~\ref{fig:type-checking}.

@@

\[
\stk{
\data \FMFreeMonad : \PITEL{\V{F}}{\Set \To \Set} 
                     \PI{\V{X}}{\Set} 
                     \Set 
\where \\
\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
    \FMVar           & X \To \FMFreeMonad\: F\: X                            \\
    \FMComposite     & F (\FMFreeMonad\: F\: X) \To \FMFreeMonad\: F\: X    
\end{array}
}
\]

@@

\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{D}{X}})}) \To
           \descop{\toDesc{\FreeMonad{D}{X}}}{\Mu{(\toDesc{\FreeMonad{D}{Y}})}} \To
           \Mu{(\toDesc{\FreeMonad{D}{Y}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\:\mapsto\:}l}
\apply\: D\: X\: Y\: \sigma\: \pair{\DVar{}}{x}{}   & \sigma\: x                   \\
\apply\: D\: X\: Y\: \sigma\: \pair{\etag{t}}{ys}{} & \Con{\pair{\etag{t}}{ys}{}}
\end{array}
}\]

@@

So, secondly, we have equipped this new data-type with its inherent
structure: the \bind and \return operations. We have been able to
build them as generic functions. Here, we rely on our ability to
compute over descriptions, thanks to the induction principle.

>>>>

Let us examine the consequences of our act. First of all, we discover
that the type theory does not actually need to be extended with the
type former $\EnumU$, the constructors $\NilE$ and $\ConsE$. We have
just defined $\EnumU$ above, by fix-point over the signature functor denoted
by $\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and $\ListCons$
constructors of lists.

@@

We shall now fulfil our promises for $\Desc$. We are going to
implement the codes of the universe of descriptions. Interestingly, in
and by itself, the system of codes, $\Desc$, is nothing but a data-type. We
are in the same situation than with $\EnumU$: we ought to be able to describe
the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
first-class citizen, born with the standard, generic equipment of
data-types.

@@

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT{[ \DUnit, \DSigma{}{}, \DIndx{} ]})  \\
                           & \left[\begin{array}{l}
                                   \DUnit                                \\
                                   \DSigma{\Set}{(\LAM{\V{S}} ???)}      \\
                                   \DIndx{\DUnit}                        \\
                                   \end{array}
                             \right]
\end{array}
}\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \ldots          & \:\:\ldots \\
    \DHindx         & \PI{H}{\Set} \Desc \To \Desc
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\ldots                        &  \ldots \\
\descop{\DHindx{H}{D}}{X}     &  \TIMES{(H \To X)}{\descop{D}{X}}
\end{array}
}
\end{array}
}\]

@@

@@

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT{[ \DUnit, \DSigma{}{}, \DIndx{}, \DHindx{}{} ]}) \\
                           & \left[\begin{array}{l}
                                   \DUnit                                            \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{S}{\DUnit})}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set}{(\LAM{\V{H}} \DIndx{\DUnit})}
                                   \end{array}
                             \right]
\end{array}
}\]

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $f$ is pushed into the type
$\EnumT{e} \To \Desc$. This corresponds to a finite function
definition, as formalised in Figure~\ref{fig:type-checking}.

@@

\[
\stk{
\data \FMFreeMonad : \PITEL{\V{F}}{\Set \To \Set} 
                     \PI{\V{X}}{\Set} 
                     \Set 
\where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FMVar           & X \To \FMFreeMonad\: F\: X                            \\
    \FMComposite     & F (\FMFreeMonad\: F\: X) \To \FMFreeMonad\: F\: X    
\end{array}
}
\]

@@

\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{D}{X}})}) \To
           \descop{\toDesc{\FreeMonad{D}{X}}}{\Mu{(\toDesc{\FreeMonad{D}{Y}})}} \To
           \Mu{(\toDesc{\FreeMonad{D}{Y}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: D\: X\: Y\: \sigma\: \pair{\DVar{}}{x}{}   & \sigma\: x                   \\
\apply\: D\: X\: Y\: \sigma\: \pair{\etag{t}}{ys}{} & \Con{\pair{\etag{t}}{ys}{}}
\end{array}
}\]

@@

So, secondly, we have equipped this new data-type with its inherent
structure: the \bind and \return operations. We have been able to
build them as generic functions. Here, we rely on our ability to
compute over descriptions.
